Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-07-29 Paper/app.tex
..

The grad-free approach discussed earlier in~\ref{sub:gradfree} was used to compute the
active subspace. Using the iterative procedure, the convergence of the eigenvectors
was examined by tracking $\delta w_{1,j}^{(i)}$. In Figure~\ref{fig:conv_app} (left),
components of the converged eigenvector based on 50 samples using $\tau$ = 0.2 are plotted.
To illustrate convergence, we compare them with components of the eigenvector obtained
using 100 samples (twice as many). The variation of maximum $\delta w_{1,j}^{(i)}$ between
successive iterations during the course of the iterative procedure is illustrated in 
Figure~\ref{fig:conv_app} (right). 
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv10}
\caption{Left: An illustrative comparison of individual components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  $\delta w_{1,j}^{(r)}$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:conv_app}
\end{center}
\end{figure}
%

>>>>

The grad-free approach discussed earlier in~\ref{sub:gradfree} was used to compute the
active subspace. Using the iterative procedure, the convergence of the eigenvectors
was examined by tracking $\delta w_{1,j}^{(i)}$. In Figure~\ref{fig:conv_app} (left),
components of the converged eigenvector based on 50 samples using $\tau$ = 0.2 are plotted.
To illustrate convergence, we compare them with components of the eigenvector obtained
using 100 samples (twice as many). The variation of maximum $\delta w_{1,j}^{(i)}$ between
successive iterations during the course of the iterative procedure is illustrated in 
Figure~\ref{fig:conv_app} (right). 
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv10}
\caption{Left: An illustrative comparison of individual components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  $\delta w_{1,j}^{(r)}$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:conv_app}
\end{center}
\end{figure}
%
The dominant eigenvector is observed to converge with only 50 samples in the 33-dimensional space which is remarkably efficient.

In Figure~\ref{fig:hd}, we illustrate the resulting eigenvalue spectrum and the sufficient summary 
plot.
%
\begin{figure}[htbp]
 \begin{center}
   \includegraphics[width=0.45\textwidth]{./Figures/eig_33D}
   \includegraphics[width=0.45\textwidth]{./Figures/ssp_33D}
\caption{Left: Eigenvalue spectrum of $\hat{\mat{C}}$. Right: Sufficient summary plot indicating the
existence of a potential 1-dimensional active subspace.} 
\label{fig:hd}
\end{center}
\end{figure}
%
The second eigenvalue is observed to be roughly four orders of magnitude smaller than the first 
indicating the existence of a 1-dimensional active subspace. This is further confirmed by the sufficient
summary plot which shows that the variability in $G$ can be approximated using a 1-dimensional regression
fit which could be used its a surrogate. We will verify the accuracy of the surrogate later in~\ref{sub:verify}.

Components of the dominant eigenvector in Figure~\ref{fig:conv_app} (left) are utilized for estimating the
activity scores, defined earlier in~\ref{eq:as}. As discussed earlier, the activity scores can be regarded as a 
measure for global sensitivity of the uncertain parameters. The sensitivity of ignition delay towards the 
33 uncertain parameters considered in this problem is plotted below in Figure~\ref{fig:as_33D}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/as_A_33D}
    \includegraphics[width=0.45\textwidth]{./Figures/as_E_33D}
\caption{A bar-graph illustrating individual activity scores for the uncertain $A_i$' (left) and $E_{a,i}$'s (right).}
\label{fig:as_33D}
\end{center}
\end{figure}
%
From the above plot, it can be inferred that the ignition delay is predominantly sensitivity towards $A_1$, $A_9$, and
$E_{a,1}$. Furthermore, it is moderately sensitivity towards $A_{15}$ and $A_{17}$ while sensitivity towards remaining
inputs is found to be low or negligible. Note that the sensitivity results for the pre-exponents is found to be consistent
with the 19-dimensional case presented in~\ref{sec:method}. This observation can also be exploited for reducing the 
dimensionality of the problem from 33 to 3-5 depending upon the required level of accuracy in terms of characterizing 
the uncertainty in the model output. The sensitivity-driven dimension reduction for uncertainty quantification has been
discussed in our earlier effort~\cite{Vohra:2018}. In this work, however, we focus on dimension reduction using the active 
subspace approach which is shown to yield even greater scope for dimension reduction i.e. from 33 to 1. We verify
the accuracy of the 1-d regression fit to $G(\bm{w}_1^\top\bm{\xi})$ in the following section.

\subsection{Surrogate Verification}
\label{sub:verify}

The 1-dimensional surrogate ($\mathcal{Y}$) is investigated for its accuracy as well as the ability to capture the 
uncertainty in the
model output in two ways. Firstly, we estimate the relative L-2 norm of the discrepancy~($\varepsilon_d$)
between estimates of ignition delay in the case of H$_2$/O$_2$ reaction, obtained using the 
model output and the surrogate in the following equation:
%
\be
\varepsilon_d = \frac{\|f(\bm{\theta}) - \mathcal{Y}(\bm{\xi})\|}{\|f(\bm{\theta})\|}
\ee
%
The relative norm, $\varepsilon_d$ was estimated to be 8.92$\times10^{-2}$ based on a validation test data
comprising 1000 model evaluations. Hence, in the norm-sense, it can be said that the 1-dimensional surrogate is 
reasonably accurate.

Secondly, we verify the accuracy of the surrogate in a probabilistic setting. In particular, we compare a histogram
plot of the model evaluations for ignition delay in the validation set with a PDF of the same using the 1-dimensional
surrogate as shown in Figure~\ref{fig:pdf_33D}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/pdf_comp_id_1e4}
\caption{A comparison of the histogram plot based on model evaluations and the PDF based on surrogate
predictions for ignition delay. The two plots are based on samples in the validation set.}
\label{fig:pdf_33D}
\end{center}
\end{figure}
%
In the above comparison, it is observed that the peak of the PDF generated using the 1-dimensional surrogate in the
active subspace is consistent with the modal estimate based on the histogram. Moreover, the spread in the ignition
 delay is reasonably captured by the PDF. Hence, the surrogate is also verified in a probabilistic sense.

### 2018-07-29 Paper/assa.tex
..

Consider a function $y = f(\xi_1, \xi_2, \ldots, \xi_\Np)$ where
$\xi_i$ are independent are either uniformly distributed or
distributed according to a Boltzman probability distribution.  
It was shown in~\cite{Lamboni:2013} that the total Sobol' 
index $T_i(f)$ can be bounded in terms of $\nu_i(f)$:
\begin{equation}\label{equ:sobol_bound}
T_i(f) \leq \frac{C_i}{\V(f)}\nu_i(f), \quad i = 1, \ldots, \Np,
\end{equation}
where for each $i$, $C_i$ is an appropriate \emph{Poincar\'{e}} constant.
This provides a strong theoretical basis for using DGSMs to identify 
unimportant inputs. Moreover, 
in~\cite{Vohra:2018} we introduced the normalized screening metric
\[
   \tilde{\nu}_i(f) = \frac{C_i \nu_i(f)}{\sum_{i=1}^\Np C_i \nu_i(f)},
\]
for detecting unimportant input parameters. 
Note that if $\xi_i$ are iid, the $C_i$'s will cancel and the normalized screening
metric reduces to 
\[
    \tilde{\nu}_i(f) = \frac{\nu_i(f)}{\sum_{i=1}^\Np \nu_i(f)} 
      = \frac{\vec{e}_i^T \mat{C} \vec{e}_i}{\trace(\mat{C})} 
      = \frac{\sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2}{\sum_{k = 1}^\Np \lambda_k}.
\]
The following result is a straightforward generalization of
the result from~\cite{Diaz:2016}, and 
links activity scores and Sobol' indices.
\begin{proposition}
For $1 \leq r \leq \Np$, and $i = 1, \ldots, \Np$, we have
\[
       T_i(f) \leq \frac{C_i}{\V(f)}(\alpha_i(f; r) + \lambda_{r+1}).
\]
\end{proposition}
\begin{proof}
The result follows immediately from Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}.
\end{proof}
We can also bound the normalized DGSMs using activity scores; it is easy to see  
\[
\tilde{\nu}_i(f) \leq 
\frac{ C_i \big(\alpha_i(f; r) + \lambda_{r+1}\big)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}
=\frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)} + \kappa_i \lambda_{r+1}, 
\]
whith $\kappa_i = C_i / (\sum_i C_i \alpha_i(f; r))$. 
In the case where where $\lambda_{r+1} \approx 0$, 
this motivates definition of
normalized activity scores
\[
   \tilde{\alpha}_i(f; r) =  \frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}.
\]

>>>>

Consider a function $y = f(\xi_1, \xi_2, \ldots, \xi_\Np)$ where
$\xi_i$ are independent are either uniformly distributed or
distributed according to a Boltzmann probability distribution.  
It was shown in~\cite{Lamboni:2013} that the total Sobol' 
index $T_i(f)$ can be bounded in terms of $\nu_i(f)$:
\begin{equation}\label{equ:sobol_bound}
T_i(f) \leq \frac{C_i}{\V(f)}\nu_i(f), \quad i = 1, \ldots, \Np,
\end{equation}
where for each $i$, $C_i$ is an appropriate \emph{Poincar\'{e}} constant.
This provides a strong theoretical basis for using DGSMs to identify 
unimportant inputs. Moreover, 
in~\cite{Vohra:2018} we introduced the normalized screening metric
\[
   \tilde{\nu}_i(f) = \frac{C_i \nu_i(f)}{\sum_{i=1}^\Np C_i \nu_i(f)},
\]
for detecting unimportant input parameters. 
Note that if $\xi_i$ are iid, the $C_i$'s will cancel and the normalized screening
metric reduces to 
\[
    \tilde{\nu}_i(f) = \frac{\nu_i(f)}{\sum_{i=1}^\Np \nu_i(f)} 
      = \frac{\vec{e}_i^T \mat{C} \vec{e}_i}{\trace(\mat{C})} 
      = \frac{\sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2}{\sum_{k = 1}^\Np \lambda_k}.
\]
The following result is a straightforward generalization of
the result from~\cite{Diaz:2016}, and 
links activity scores and Sobol' indices.
\begin{proposition}
For $1 \leq r \leq \Np$, and $i = 1, \ldots, \Np$, we have
\[
       T_i(f) \leq \frac{C_i}{\V(f)}(\alpha_i(f; r) + \lambda_{r+1}).
\]
\end{proposition}
\begin{proof}
The result follows immediately from Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}.
\end{proof}
We can also bound the normalized DGSMs using activity scores; it is easy to see  
\[
\tilde{\nu}_i(f) \leq 
\frac{ C_i \big(\alpha_i(f; r) + \lambda_{r+1}\big)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}
=\frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)} + \kappa_i \lambda_{r+1}, 
\]
with $\kappa_i = C_i / (\sum_i C_i \alpha_i(f; r))$. 
In the case where where $\lambda_{r+1} \approx 0$, 
this motivates definition of
normalized activity scores
\[
   \tilde{\alpha}_i(f; r) =  \frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}.
\]
